Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-12 paper/main.tex
formatting sections 4 and 5

%SHORT
% \lstset{basicstyle=\tt\scriptsize}
% \abovedisplayskip=3pt
% \belowdisplayskip=3pt
% \abovedisplayshortskip=3pt
% \belowdisplayshortskip=3pt
%SHORT

@@

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
\itemskip
\itemskip
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
\vspace{-1.5ex}
$$
then we get the following familiar looking rule
\itemskip
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
\vspace{-1.5ex}
$$
%
This rule already shows a departure from classical techniques that are
based on ranking functions.  Reasoning with two potential functions
promotes compositional reasoning by considering the sequencing of
program operations.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
was derived on $S_1$ only.
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\begin{figure*}
 \setlength{\progwidth}{.22\linewidth}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y)
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y=y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x=x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[y, m]|$
\\[.4\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

@@

\begin{figure*}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{7cm}
    \begin{center}
   \begin{lstlisting}[]
void c_down (int x,int y) {
  if (x>y) {tick(1); c_up(x-1,y);}
}
void c_up (int x, int y) {
  if (y+1<x) {tick(1); c_down(x,y+2);}
}
   \end{lstlisting}
\vspace{-2.5ex}
$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
\\[.4\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5cm}
    \begin{center}
   \begin{lstlisting}
for (; l>=8; l-=8)
  /* process one block */
  tick(N);
for (; l>0; l--)
  /* save leftovers */
  tick(1);
   \end{lstlisting}
\vspace{-2.5ex}
$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
\\[.4\baselineskip]
      {\bf t61}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5.5cm}
    \begin{center}
   \begin{lstlisting}

@@

\ifshort{
\begin{figure*}
\small
\begin{mathpar}
\Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
{ \Gamma \models y \ge 0 \!
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx} \!
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\\ \Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ \!\! \Gamma_1 \models \Gamma_2
\\\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\\ \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q' \succeq_\Gamma Q }

@@

\section{Derivation System}

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$

These judgements are similar to classic Hoare triples in the presence
of the early exit statements \code{break} and \code{return}.  The part
$(\Gamma_B, Q_B)$ is the postcondition used when breaking out of
a loop.  For example, it is set upon entry of looping statements by
the rule {\sc Q:Loop} and used before \code{break} statements by the
rule {\sc Q:Break}.  Similarly $(\Gamma_R, Q_R)$ is the postcondition
used before returning from a function call.  The part $(\Gamma, Q) \vdash
S \dashv (\Gamma', Q')$ of the judgement can be seen as a regular
Hoare triple.  All assertions are split into two parts, the logical
part and the quantitative part.  The quantitative part $Q$ represents a
potential function as a collection of non-negative numbers $q_i$ indexed
by the index set $I$.  The logical part $\Gamma$ is left abstract but
is enforced by our derivation system to respect classic Hoare logic
constraints.

@@

As a convention, if $Q$ and $Q'$ are quantitative annotations
we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
The notation $Q \pm n$ used in many rules defines a new context $Q'$
such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
In all the rules, we have the implicit side condition that all rational
coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
and leaves the latter undefined at some index $i$ we will implicitly
assume that $q'_i = q_i$.

We describe the automatic amortized analysis for a subset of
expressions of Clight.  Assignments are restricted to the form $x \gets
y$ or $x \gets x \pm y$.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This is achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
a[y]$ are simply handled by assigning zero potential to coefficients
like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.

\paragraph{Derivation Rules.}

The rules of our derivation system must serve two purposes.  They must
\emph{attach} potential to certain program variable intervals and use
this potential, when it is allowed, to \emph{pay} for resource consuming
operations.  These two purposes are illustrated on the {\sc Q:Skip}
rule.  This rule reuses its precondition as postcondition, it is
explained by two facts: first, no resource is consumed by the skip
operation, thus no potential has to be used to pay for the evaluation;
second, the program state is not changed by
the execution of a skip statement, thus all the potential available
before the execution of the skip statement is still available after.
These two observations would also justify a rule with the side-condition
$Q \ge Q'$, because some potential is always safely lost.  For the sake
of orthogonality, it is not how we present the rule and this form of
weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
described later.

@@

Most interesting are the rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
which describe how the potential is distributed after a size change of
a variable.
\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
$x \gets x+y$, it applies when we can deduce from the logical context
$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
can be applied for decrements and when $y$ is not positive.}
The rules are all equivalent in the case where $y=0$.
Finally, the rule {\sc Q:Inc} can be applied
if we cannot deduce any information about $y$.  In the implementation,
the rules for increments and decrements are combined into one syntax
directed rule where the specialized rules take priority over {\sc Q:Inc}.

\iffull{To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The others follow the
same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule ensures this
only for $v \not\in \mathcal U$ because we know that $x \le v$
otherwise, and thus $|\inter v x| = 0$.

@@

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  \iffull{We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the analysis
efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
deals also with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.

>>>>

%SHORT
\ifx\fullversion\undefined
\abovedisplayskip=3pt
\belowdisplayskip=3pt
\abovedisplayshortskip=3pt
\belowdisplayshortskip=3pt
\fi
%SHORT

@@

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
$$
%
This rule already shows a departure from classical techniques that are
based on ranking functions.  Reasoning with two potential functions
promotes compositional reasoning by considering the sequencing of
program operations.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
was derived on $S_1$ only.
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\begin{figure*}
 \setlength{\progwidth}{.22\linewidth}
  \centering
\vspace{-.6cm}
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y)
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y=y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x=x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}
\vspace{-2.5ex}
$|[x, n]| + |[y, m]|$
\\[.4\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

@@

\begin{figure*}
  \centering
\vspace{-1.8cm}
\hspace{-0.4cm}
  \begin{minipage}[b]{7cm}
    \begin{center}
   \begin{lstlisting}[]
void c_down (int x,int y) {
  if (x>y) {tick(1); c_up(x-1,y);}
}
void c_up (int x, int y) {
  if (y+1<x) {tick(1); c_down(x,y+2);}
}
   \end{lstlisting}
\vspace{-2.5ex}
$0.33 + 0.67 |[y,x]|\;\;$ (\code{c\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{c\_up(x,y)})
\\[.4\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5cm}
    \begin{center}
   \begin{lstlisting}
for (; l>=8; l-=8)
  /* process one block */
  tick(N);
for (; l>0; l--)
  /* save leftovers */
  tick(1);
   \end{lstlisting}
\vspace{-2.5ex}
$\;\;\;\;\;\;\;\;\;\;\;\;\;\frac{N}{8} |[0,l]|\;\;$ if $N \geq 8$\\
$7\frac{8-N}{8} + \frac{N}{8}|[0,l]|\;\;$ if $N < 8$
\\[.4\baselineskip]
      {\bf t61}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{5.5cm}
    \begin{center}
   \begin{lstlisting}

@@

\ifshort{
\begin{figure*}
\small
\vspace{-.3cm}
\begin{mathpar}
\Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
{ \Gamma \models y \ge 0 \!
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx} \!
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\\ \Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ \!\! \Gamma_1 \models \Gamma_2
\\\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\\ \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q' \succeq_\Gamma Q }

@@

\sectskip
\section{Derivation System}

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
%
The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
can be seen as a regular Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resource units are available then the
evaluation does not run out of resources and---if the execution
terminates in state $\state'$---there are at least $Q(\state')$
resources left and $\Gamma'$ holds for $\state'$.

The judgement is a bit more involved since we have to take into
account the early exit statements \code{break} and \code{return}.
This is similar to classical Hoare triples the presence of non-linear
control flow. In the judgement, $(\Gamma_B, Q_B)$ is the postcondition
that holds when breaking out of a loop using \code{break}.  % For example, it is set upon entry
% of looping statements by the rule {\sc Q:Loop} and used before
% \code{break} statements by the rule {\sc Q:Break}.  
Similarly, $(\Gamma_R, Q_R)$ is the postcondition that holds when
returning from a function call with \code{return}. (This becomes clear
in the rules.)

@@

As a convention, if $Q$ and $Q'$ are quantitative annotations
we assume that $Q = (q_i)_{i\in I}$ and $Q' = (q'_i)_{i \in I}$.
The notation $Q \pm n$ used in many rules defines a new context $Q'$
such that $q'_0 = q_0 \pm n$ and $\forall i \neq 0 .\, q'_i = q_i$.
In all the rules, we have the implicit side condition that all rational
coefficients are non-negative.  Finally, if a rule mentions $Q$ and $Q'$
and leaves the latter undefined at some index $i$ we assume that $q'_i = q_i$.

We describe the automatic amortized analysis for a subset of
expressions of Clight.  Assignments are restricted to the form $x \gets
y$ or $x \gets x \pm y$.  In the implementation, a Clight
program is converted into this form prior to analysis without changing
the resource cost.  This is achieved by using a series of \emph{cost-free
  assignments} that do not result in additional cost in the
semantics.  Non-linear operations such as $x \gets z*y$ or $x \gets
a[y]$ are handled by assigning $0$ to coefficients
like $q_{xa}$ and $q_{ax}$ that contain $x$ after the assignment.

\paraskip
\paragraph{Derivation Rules.}

The rules of our derivation system must serve two purposes.  They must
\emph{attach} potential to certain program variable intervals and use
this potential, when it is allowed, to \emph{pay} for resource consuming
operations.  These two purposes are illustrated on the {\sc Q:Skip}
rule.  This rule reuses its precondition as postcondition, it is
explained by two facts: first, no resource is consumed by the skip
operation, thus no potential has to be used to pay for the evaluation;
second, the program state is not changed by
the execution of a skip statement, thus all the potential available
before the execution of the skip statement is still available after.
\iffull{
These two observations would also justify a rule with the side-condition
$Q \ge Q'$, because some potential is always safely lost.  For the sake
of orthogonality, it is not how we present the rule and this form of
weakening is in fact expressed as the stand-alone rule {\sc Q:Weak}
described later.}{}

@@

The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
$x \gets x+y$, it applies when we can deduce from the logical context
$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
can be applied for decrements and when $y$ is not positive.}
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot deduce any information about $y$.  In the implementation,
the rules for increments and decrements are combined into one syntax
directed rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}

\iffull{To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The others follow the
same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule ensures this
only for $v \not\in \mathcal U$ because $x \le v$
otherwise, and thus $|\inter v x| = 0$.

@@

The rule {\sc Q:Weak} is not syntax directed.  In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition.  \iffull{We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the analysis
efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}.  Here, $Q$ and $Q'$ are potential
annotations and $\Gamma$ is a logical context.  The rule {\sc Relax}
deals also with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$\iffull{ without gaining potential.}{.}
